---
bibliography: references.bib
---

# Data {#sec-data}

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(readr)
library(ozmaps)
library(sf)
library(zoo)
library(lubridate)
library(gridExtra)
library(ggpubr)
```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
lonlat <- data.frame(station = c("Richmond UWS Hawkesbury",
                                 "Kurrajong Heights",
                                 "Sydney",
                                 "Newcastle"),
                     longitude = c(150.7477,
                                   150.6338,
                                   151.216,
                                   151.75),
                     latitude = c(-33.6165,
                                  -33.5343,
                                  -33.8662,
                                  -32.9185)) %>%
  as_tibble()

aus <- ozmap("states")
nsw <- aus %>%
  filter(NAME == "New South Wales")
```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
psoi <- ggplot(data = nsw) +
  geom_sf() +
  geom_point(data = lonlat,
             aes(x = longitude, y = latitude, color = station)) +
  xlim(150,152.5) +
  ylim(-34.5,-32.5) +
  labs(x = "Longitude",
       y = "Latitude",
       colour = "Station") +
  theme_bw()
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#| label: fig-map
#| fig-cap: "Map of the Australian Eastern Seaboard, showing the location of the stations analysed."
psoi
```

This paper focuses on two different stations in the Hawkesbury River region, with Sydney and Newcastle as comparison points (@fig-map) to analyse the frequency and drivers of extreme rainfall events in the region relative to the nearest population hubs.

The data sources for this research are shown in @tbl-data.

| Data                                  | Source                                                                                     |
|----------------------------------|--------------------------------------|
| New South Wales Station Rainfall Data | Global Historical Climatology Network [@chamberlain2019]                                   |
| New South Wales Gridded Rainfall Data | National Oceanic and Atmospheric Administration Physical Sciences Laboratory [@kalnay1996] |
| ENSO & IOD Data                       | Bureau of Meteorology [@climate]                                                           |
| Southern Annular Mode Data            | Climate Data Guide [@marshall2003]                                                         |
| Atmospheric Pressure & Wind Data      | Climate Data Store [@hersbach2020]                                                         |

: Datasets and sources for the research. {#tbl-data}

The precipitation data has been obtained using the `rnoaa` package, pulling weather station data for the Hawkesbury River region from the Global Historical Climatology Network's daily data (@tbl-data). The Hawkesbury stations analysed were chosen with consideration given to length of the series, quality of the data, and the degree of data missingness.

## Data Quality Considerations

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
clean <- read_csv("cleanareasofinterestv2.csv")
# Transformation of Data
stations <- clean %>%
  filter(!is.na(prcp))

v <- unlist(stations$prcp) %>%
  as.integer()

stations <- stations %>%
  select(-prcp) %>%
  mutate(prcp = v) %>%
  mutate(mm = prcp / 10)

stations.nonzero <- stations %>%
  filter(prcp > 0)

## Nonzero Series
richmond.nonzero <- stations.nonzero %>%
  filter(location=="Richmond UWS Hawkesbury") %>%
  filter(date >= "1950-01-01" & date <= "2023-01-30")
kurrajong.nonzero <- stations.nonzero %>%
  filter(location=="Kurrajong Heights") %>%
  filter(date >= "1950-01-01" & date <= "2023-01-30")
sydney.nonzero <- stations.nonzero %>%
  filter(location=="Sydney Botanic Gardens") %>%
  filter(date >= "1950-01-01" & date <= "2023-01-30")
newcastle.nonzero <- stations.nonzero %>%
  filter(location=="Newcastle Nobbys Signal Station") %>%
  filter(date >= "1950-01-01" & date <= "2023-01-30")
```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
# Sunday / Monday Accumulation
## Day of Week Analysis
day <- clean %>%
  select(-year) %>%
  mutate(day = wday(date, label = TRUE)) %>%
  filter(!is.na(prcp))

day.agg.kurr <- day %>%
  filter(location == "Kurrajong Heights") %>%
  filter(date <= "2023-01-30") %>%
  mutate(mm = prcp / 10) %>%
  select(mm, day)

day.agg.kurr <- day.agg.kurr %>%
  group_by(day) %>%
  summarise(tot = sum(mm))

sm <- ggplot(data = day.agg.kurr,
       aes(x = day, y = tot, fill = day)) +
  geom_bar(stat = "identity") +
  scale_fill_hue(c = 50) +
  labs(x = "Day of the Week",
       y = "Accumulation (mm)",
       fill = "Day") +
  theme_bw()

```

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
## Missingness
numobs_keyid <- clean %>%
  filter(!is.na(prcp)) %>%
  group_by(id, year) %>%
  count(prcp >= 0) %>%
  ungroup()

numobs_keyid <- numobs_keyid %>%
  filter(year <= 2022) %>%
  filter(year >= 1950)

numobs_keyid <- numobs_keyid %>%
  mutate(location = ifelse(id=="ASN00063043", "Kurrajong Heights",
                    ifelse(id=="ASN00067021", "Richmond UWS Hawkesbury",
                    ifelse(id == "ASN00066006", "Sydney Botanic Gardens",       
                    ifelse(id == "ASN00061055", "Newcastle Nobbys Signal Station", NA))))) %>%
  mutate(missing = 365 - numobs_keyid$n) %>%
  mutate(missing = ifelse(missing == -1, 0, missing)) %>%
  na.omit()

miss <- ggplot() +
  geom_line(data = numobs_keyid,
            aes(x = year, y = missing, color = location)) +
  labs(x = "Year",
       y = "Missing Observations",
       labs = "Data Missingness",
       color = "Location") +
  theme_bw()
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#| label: fig-data
#| fig-cap: "Examples of data quality Considerations, showing Sunday-Monday accumulations and data missingness at each location."
ggarrange(sm, miss,
          nrow = 2)
```

Data quality issues may arise where there are issues with the recording station, with rainfall totals needing to be verified by the presence of ground truth. Sunday/Monday accumulations [@viney2004] (@fig-data) are a reporting anomaly where historically, rainfall observations may not have been taken on the Sunday and instead aggregated with the Monday rainfall, leading to a daily distribution of rainfall that has less rain occurring on Sunday's and greater amounts on Monday's. This phenomenon can lead to multiday accumulated rainfall totals being recorded as an extreme event, despite not being so, on the day the observation is recorded.

Data missingness exists where observations are missing, either due to the complete absence of observation or due to being quality flagged (@fig-data). Missingness is detrimental in an extremes setting where data points are limited. The extreme event itself may be the cause of the missingness, as seen in the 2022 Lismore floods, where flood waters washed away the entire network of gauges. When the extreme weather event is the cause of missingness, the data of interest for extremes analysis is lost, potentially requiring adjustments for any statistical analysis performed.

This research considers daily observations from the 1st of January 1950 onwards in an attempt to mitigate some of the data quality issues caused by historical untagged accumulations and data missingness. For days on which precipitation totals were missing, gridded data encompassing the station was used. The use of gridded data is an appropriate approximation for days where rainfall data is missing, as it interpolates daily rainfall totals for a small geographic region, providing an estimate reflective of reality for a station.

## Atmospheric Drivers Impacting the Australian Climate

```{r, include = FALSE, echo = FALSE, warning = FALSE, message = FALSE}
richmond <- read_csv("richmond.csv")
lanina <- read_csv("lanina.csv")
colours4 <- c("La Nina" = "blue",
              "El Nino" = "red")
colours2 <- c("La Nina" = "skyblue")
soi.plot <- richmond %>%
  ggplot() +
  geom_rect(data = lanina, aes(xmin = xmin, xmax = xmax, ymin = ymin,
                  ymax = ymax, fill = "La Nina"), alpha = 0.5) +
  scale_fill_manual(values = colours2) +
  geom_line(aes(date, soi)) +
  geom_hline(aes(yintercept = 7, colour = "La Nina")) +
  geom_hline(aes(yintercept = -7, colour = "El Nino")) +
  scale_colour_manual(values = colours4) +
  theme_bw() +
  ylab("SOI") +
  xlab("Date") +
  labs(fill = "La Nina Event",
       colour = "ENSO")

colours <- c("Positive" = "red",
             "Negative" = "blue")
dmi.plot <- richmond %>%
  ggplot() +
  geom_line(aes(date, dmi)) +
  geom_hline(aes(yintercept = 0.4, colour = "Positive")) +
  geom_hline(aes(yintercept = -0.4, colour = "Negative")) +
  scale_colour_manual(values = colours) +
  theme_bw() +
  labs(y = "DMI", x = "Date", colour = "IOD")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#| label: fig-driver
#| fig-cap: "Time Series of SOI (ENSO) and DMI (IOD), showing the transition between SST anomaly phases since 1950."
grid.arrange(soi.plot, dmi.plot)
```

The ENSO and IOD data obtained from the Bureau is used to incorporate large-scale climate driver covariates in the modelling of extremes arrivals. The Southern Oscillation Index (SOI) serves as a proxy for the phase of ENSO and the DMI (Dipole Mode Index) index as proxy for the phase of IOD. Data for the Southern Annular Mode is obtained from the Climate Data Guide [@marshall2003].

Analysis of the relationship between localised atmospheric conditions and the temporal clustering of extremes is conducted using mean sea level pressure (MSLP) and wind speed/direction data obtained using ERA5 Reanalysis data from the Climate Data Store [@hersbach2020]. Low pressure systems cause lower temperatures and higher rainfall totals, with impacts exacerbated if winds are easterly and the system remains stationary over an area for a period of time. Unlike large-scale atmospheric drivers such as ENSO or IOD, atmospheric pressure and wind speed/direction are highly localised and provide more location dependent information. The observed daily pressure and wind values used for analysis are calculated using a simple average based upon the values observed at 3am, 9am, 3pm and 9pm each day.

ENSO is the predominant large-scale climate driver impacting rainfall variability in an Australian context [@king2015], and is an important consideration when analysing the temporal clustering of rainfall extremes in the region. ENSO refers to a phenomenon in the Pacific Ocean where SSTs periodically warm and cool and Pacific trade winds strengthen and weaken, impacting the global climate [@wang2017]. The El Niño phase is associated with a warming of SSTs in the Pacific Ocean, stronger trade winds and results in higher temperatures and less rainfall in Eastern Australia. Alternatively, the La Niña phase is associated with the cooling of SSTs in the Pacific Ocean, weaker trade winds and results in lower temperatures and more rainfall in Eastern Australia.

SOI acts as a proxy for ENSO's current phase and is calculated using differences in pressure between Darwin and Tahiti. Sustained SOI values above +7 indicate a La Niña phase and sustained values below -7 indicate an El Niño phase (@fig-driver).

IOD is an SST anomaly occurring in the Indian Ocean, and is a key driver of rainfall and temperature throughout Southern Australia [@ummenhofer2011], regulating moisture flow to Australia. IOD consists of three phases, positive, neutral and negative, with a negative IOD driving cooler temperatures and increases to rainfall in southern Australia. The DMI index is calculated by taking the SST difference between the tropical western Indian Ocean and the eastern Indian Ocean. Sustained DMI values above +0.4 indicate a positive phase and sustained values below -0.4 indicate a negative phase (@fig-driver).

The Southern Annular Mode (SAM) affects rainfall and temperatures in southern Australia, shifting westerly winds in the Southern Ocean either north or south. These winds are typically strong and have the ability to carry rain, resulting in thunderstorms and cooler conditions when impacting Australia. A positive SAM phase occurs where westerly winds in the Southern Ocean are further south, resulting in increased rainfall and cooler temperatures for Eastern Australia. A negative event occurs when winds are further north and the impact on temperature and rainfall is seasonal. The SAM index is calculated by taking the pressure differences between stations at a latitude of 40 degrees south and stations at a latitude of 65 degrees south [@gong1999].

When these large-scale atmospheric drivers interact, their impacts are magnified, increasing the risk of extreme weather events. Two examples of these interaction effects are the 2022 Eastern Australia floods which occurred during a La Niña and negative IOD period; and the Ash Wednesday bushfires which occurred during an El Niño and positive IOD period.

Monthly averages of the SOI, IOD and SAM indexes are used for analysis.
